\chapter*{Introduction} % chapter* je necislovana kapitola
\addcontentsline{toc}{chapter}{Introduction} % rucne pridanie do obsahu
\markboth{Introduction}{Introduction} % vyriesenie hlaviciek

In the first chapter, we introduce the field of succinct data structures.
We explain the goal of these data structures and describe scenarios when
they are very competitive and useful but we also discuss their drawbacks.
We introduce bit vector, one of the simplest succinct data structures, and
demonstrate some of its useful applications. In the end, we demonstrate how
bit vector can be used inside of the text indexing data structure FM-index.

In the second chapter, we show possible implementations of methods $\rank$ and
$\select$. We discuss what are the theoretically optimal solutions but also
what are their practical alternatives. We then proceed to describe one of the
widely used implementations of the compressed bit vector called \textit{RRR}.

In the third chapter, we propose our own modifications to the implementation of
compressed bit vector based on RRR. We discuss theoretical aspects of these
modifications.

In the fourth chapter, we show our proposed implementation of previously devised
methods and experimentally test and evaluate them. We measure the performance
of our solution on artificial as well as real-world data. Finally, we demonstrate
how our idea can be used to get a better ratio of query time and space used by the
FM-index and thus deliver practically useful results.